Seismic parameter estimation and the Canadian crust

### 2013-11-24 paper/draftBP.tex
intro almost complete

The Lithoprobe project has provided researches with a rich dataset from which detailed cross-sections have been derived. These cross sections form a near continuous transect from the Pacific to the Atlantic oceans have and enabled many insights into crustal structure and genesis. Recent examples include Cook et al [2009] analysis of the Moho where they present the Moho as having varied and complex origins and as evolving through thermal and pressure changes, deformation and fluid migration. In another recent Lithoprobe study, Hammer and Clowes [2011] discuss broad structural characteristics of the Canadian crust. They note a remarkable consistency in orogenic style such as the similarity between geolgical structures contained within Archean cratons and their boundaries as well as those observed in Proterozoic and Phanerozoic orogens. Also noted is the striking consistency in moho depth, which is imaged as lying between 33km and 43km depth along the full 20,000km of Lithoprobe transects, despite the diversity of overlying crust. Both Cook et al. and Hammer and Clowes raise the possibility that in some situations the Moho and the base of seismic reflectivity do not represent the petrological crust-mantle boundary. A final example of a Canada wide crustal study utilizing Lithoprobe data is the Lith5.0 revised crustal model presented by Perry, Eaton and Forte [2002]. This work updates the Crust5.1 model [Mooney et al, 1998] to include the new data from Lithoprobe and introduces broad crustal features of the Canadian crust such as the relatively thin Archean Shield regions (32.5-40.0 km) as compared to the thick (>42.5km) Phanerozoic crust in the south and west of the continental craton. These studies and other Canada wide studies employ active source Lithoprobe data which records higher frequency (10-40hz) {\it P}-wave velocity whereas this study uses lower frequency (0.4-3hz) passive source scattering techniques which provides estimates of the $V_P/V_S$ ratio and images relatively larger scale structure. Studies of the Canadian crust employing a similar teleseismic receiver function method as this work include a study of the Great Slave Lake shear zone [Eaton and Hope, 2002] which utilize receiver functions, shear-wave splitting results and gravity data to characterize the shear zone. Finally, three teleseismic studies discussed in greater detail in later sections are presented in chronological order: an examination of crustal thickness and $V_P/V_S$ variation in the Grenville Orogen [Eaton et al., 2005], a structural and compositional characterization of the Superior Province [Darbyshire et al. 2006] and a crustal evolution analysis of the Hudson Bay region [Thompson et al., 2010]. This work differs from these studies in that it focuses on interregional comparisons and Canada wide structural characteristics, as does some of the Lithoprobe literature, but also on Canadian bulk crustal composition.

While there is a great diversity and body of seismic research on the Canadian crust those employing similar teleseismic methods and those encompassing the whole Canadian landmass require mention here. Recent teleseismic receiver function studies in the Candian Shield include

In the following sections we describe the data and methods that we employ to interrogate the Canadian crust. We then proceed to present the major results beginning with a comparison between our seismic parameter estimates and the estimates from three published for matching seismic stations. We then present the results of a modified approach to our parameter estimation method and follow this with a presentation of the main dataset in two sections, one focusing on crustal discontinuities and structure and the other on regional variation and characterization. Finally we conclude with a discussion of implications of the results for our understanding of the Conrad discontinuity and bulk crustal composition of the Canadian Crust.
%% Seismic studies of the Canadian continental crust predominately analyze specific geological features or regions rather than taking a comprehensive and comparative inter-regional analysis. The primary reason for this is the poor resolution afforded by the seismic networks currently and historically deployed across Canada. The Canadian continental landmass is composed of at least fifteen large geological provinces as recognized by the Geological Survey of Canada (GSC). Each of these regions is itself complex and heterogeneous and often larger than most European nations. On top of this there is poor seismic coverage, roughly one seismic station per 25,000 $km^2$. Many of these stations are clustered near areas of geologic interest such as the Cascadia Subduction zone or population and research centres such as Southern Ontario or areas of significant resource interest like the diamondiferous Great Slave Lake area. This leaves vast areas of the Canadian landmass completely unsampled. Despite these limitations, a low resolution comprehensive and comparative study is still feasible for some of the geological regions comprising the Canadian continental crust. Direct comparison between the bulk average seismic properties of geological regions and between aggregated regions and global averages provide some new insight into the crustal composition of the Canadian landmass. The accumulated dataset also affords investigations into variations of bulk crustal parameters such as Poisson's Ratio and crustal thickness with age and tectonic environment as well as some basic statistical data on interesting crustal features.

%% This paper presents a comparative tour through the dataset accumulated from processing more than a decade worth of data from all available Canadian seismic stations. It begins with a discussion of the raw data itself followed by a review of the receiver function method and a more detailed explanation of the inversion algorithms used to produce the dataset. Three previous publications utilizing similar processing schemes provide unique subsets of data to compare with the values computed in this survey for quality assurance. ...

>>>>

The Canadian continental lithosphere is composed of more than fifteen large geological provinces representing crustal formation and evolution from the early Archean to the present accretionary growth and orogenic deformation in the Western Cordillera. The Precambrian Canadian Shield and the extensive platforms of North America's interior form part of the assemblage that comprises the cratonic core, Laurentia. The Canadian Shield is itself composed of several distinct geological provinces brought together during a series of Paleoproterozoic orogens [Hoffman, 1998]. The largest of these provinces (the world's largest Archean craton) is the Superior Province. Bounded by Proterozoic orogenic belts the western and sourthern Superior is composed of roughly eastwest trending geological subprovinces that include the granite-greestones of the Abitibi Greenstone Belt, metasedimentary gneisses, high-grade gneisses and plutonic complexes assembled during the period between 3-2.6Ga [Darbyshire et al., 2006]. In the northwest of the Superior Province on the Eastern shores of the Hudson Bay is the Nuvvuagittuq greenstone belt which recent geochronology work has suggested formation times as early as ~4.28Ga making it the only known remnant of Hadean crust preserved on Earth [O'Neil et al., 2011]. To the east is the relatively younger, 1.2-0.98Ga, Grenville orogen, consisting largely of Superior and Southern Province rock reworked during Alpine-Himalayan like continental collision [Eaton et al., 2006]. To the north of the Superior Province and separated by the Paleoproterozoic Trans-Hudson Orogen are the Rae and Hearne domains which together form the Western extent of the Churchill Province. Work in the the Rae suggest a Paleoarchean and Mesoarchean basement whereas the Hearne comprises granite-and-greenstone terranes of ages ~2.7Ga [Thompson et al., 2010]. To the northwest of the Superior is the Slave Province. A relatively small cratonic fragment, the Slave is an Archean granite-greenstone block with geochronology dating between 4.0 Ga in the western Slave Province, the remnant of a larger Archean continent, to 2.6Ga in the eastern Slave, a younger paired accretionary prism and island arc [Bostock, 1998]. To the west of the Superior are the extensive Interior Platforms comprising the southern extents of the Trans-Hudson Orogen and Hearne Domain as well as the Alberta Orogen. The Interior platform is largely Archean reworked during the Paleoproterzoic with bore-hole results showing amphibolite to granulite-grade Proterozoic metamorphism [Hammer et al., 2010]. Extending west from the Interior platform the Western Cordillera is composed of amalgamated island arcs accreted after the onset of subduction during the Mesozoic. Although there is considerable variation in structure and tectonic history roughly half of the Cordillera is underlain by Proterozoic-margin crust which forms a crustal-scale décollement beneath interleaved and duplexed crust. Although there is significant orogeny in the southeast studies have shown that the crust is thin and relatively uniform in thickness with no crustal root being observed [Hammer et al., 2010]. The rich tectonic history of the Canadian continental crust affords an excellent opportunity to study crustal properties and composition with geological age. [[ THAT LAST SENTENCE WAS KILLING ME, I GAVE UP... IT ALL SOUNDED TRITE TO ME ]]

The Lithoprobe project has provided researches with a rich dataset from which detailed cross-sections have been derived. These cross sections, which form a near continuous transect from the Pacific to the Atlantic Oceans, have enabled many insights into crustal structure and genesis. Recent research utilizing lithoprobe data include the analysis of Cook et al [2009] where the Moho is argued as having varied and complex origins and as evolving through thermal and pressure changes, deformation and fluid migration. In another recent Lithoprobe study, Hammer and Clowes [2011] discuss broad structural characteristics of the Canadian crust. They note a remarkable consistency in orogenic style such as the similarity between geolgical structures contained within Archean cratons and their boundaries as well as those observed in Proterozoic and Phanerozoic orogens. Also noted is the striking consistency in moho depth, which is imaged as lying between 33km and 43km depth along the full 20,000km of Lithoprobe transects, despite the diversity of overlying crust. Both Cook et al. and Hammer and Clowes raise the possibility that in some situations the Moho and the base of seismic reflectivity do not represent the petrological crust-mantle boundary. A final example of lithoprobe work is the Lith5.0 revised crustal model presented by Perry, Eaton and Forte [2002]. This work updates the Crust5.1 model [Mooney et al, 1998] to include the new data from Lithoprobe and introduces broad crustal features of the Canadian crust such as the relatively thin Archean Shield regions (32.5-40.0 km) as compared to the thick (>42.5km) Phanerozoic crust in the south and west of the continental craton. This work shares some features with the Lithoprobe research in that it is both regionally comparitive and concerned with Canadian bulk crustal characteristics. However, the focus on crustal reflectivity patterns and crustal composition is unique to these other works. Additionally, whereas the studies and other Lithoprobe research rely on the relatively high frequency (10-40hz) {\it P}-waves as the primary imaging source this study uses lower frequency (0.4-3hz) passive source scattering techniques. Therefore Lithoprobe data represents velocity contrasts in tens to hundreds of meters and resolves {\it P}-wave velocity while the data employed in this study is sensitive to larger scale lengths (1000-2000m) and resolves the {\it P} and {\it S} wave ratio. Studies of the Canadian crust employing similar teleseismic scattering techniques as this work include a study of the Great Slave Lake shear zone [Eaton and Hope, 2002] which utilize receiver functions, shear-wave splitting results and gravity data to characterize the shear zone. Finally, three teleseismic studies discussed in greater detail in later sections are here presented in chronological order: an examination of crustal thickness and $V_P/V_S$ variation in the Grenville Orogen [Eaton et al., 2005], a structural and compositional characterization of the Superior Province [Darbyshire et al. 2006] and a crustal evolution analysis of the Hudson Bay region [Thompson et al., 2010]. This work differs from these teleseismic studies in both scale and primary focus. Whereas the teleseismic studies tend to interrogate a particular geological region or set of related regions this study looks at both differences across regions and the aggregate bulk Canadian crust. Finally, this study does not seek to define particular geological sequences but rather ... bulk crustal reflectivity, discontinuity and composition.

Studies on bulk crustal composition are better represented in the global continental crust literature. Two widely cited studies [Christensen and Mooney, 1995, Ruknick and Fountain, 1995] use refraction data to develop a velocity model which is used to constrain ratios of rock units assigned to different crustal depths. The chemical compositions of these ratios are aggregated to develop compositional estimates for the lower, mid and upper crust as well as for the entire crust. This work is different in that it employs updated rock velocity measurements [Christensen, 1996] and utilizes the reciever function method, which constrains the compositional analysis to the aggregated bulk crust. Two studies of crustal evolution use seismic parameter estimation and  by Zandt and Ammon [1995] employs teleseismic reflections to estimate Poisson's Ratio at 114 global broadband stations which are then sorted by geological age.

In the following sections we describe the data and methods that we employ to interrogate the Canadian crust. We then proceed to present the major results beginning with a comparison between our seismic parameter estimates and the estimates from three published for matching seismic stations. We then present the results of a modified approach to our parameter estimation method and follow this with a presentation of the main dataset in two sections, one focusing on crustal discontinuities and structure and the other on regional variation and characterization. Finally we conclude with a discussion of implications of the results for our understanding of the Conrad discontinuity and bulk crustal composition of the Canadian Crust.

This paper presents a comparative tour through the dataset accumulated from processing more than a decade worth of data from all available Canadian seismic stations. It begins with a discussion of the raw data itself followed by a review of the receiver function method and a more detailed explanation of the inversion algorithms used to produce the dataset. Three previous publications utilizing similar processing schemes provide unique subsets of data to compare with the values computed in this survey for quality assurance. ...
